---
title:  "Physical Activity Trends by Cancer History in BRFSS, 2018–2024"
author: "Biruk Demma"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



 This script analyzes BRFSS 2018–2024 data to:
 1. Estimate survey-weighted prevalence of physical activity
    among adults with and without a history of cancer.
 2. Evaluate temporal trends in physical activity prevalence using annual
    percentage change (APC).
 3. Estimate adjusted prevalence differences in physical activity by cancer history.
 4. Conduct a secondary subanalysis of aerobic guideline adherence in 2019 and 2023.



```{r}

#load packages
pacman::p_load( haven, dplyr, tidyr, tibble, ggplot2, survey, broom, gtsummary, knitr, grid)
```


 Read BRFSS annual files (2018–2024)

 Only required variables are imported.
 Note: variable names differ by year 
 
```{r}

# 2024

BRFSS_2024 <- read_xpt(
  "LLCP2024.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2",
                 "_AGE80","_AGE_G","EDUCA","_HLTHPL2","MARITAL","PERSDOC3",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2024)


# 2023

BRFSS_2023 <- read_xpt(
  "LLCP2023.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2", "_PA150R4",
                 "_AGE80","_AGE_G","EDUCA","_HLTHPL1","MARITAL","PERSDOC3",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2023)


# 2022

BRFSS_2022 <- read_xpt(
  "LLCP2022.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPR1","CHCOCNC1","CNCRTYP2",
                 "_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2022)


# 2021

BRFSS_2021 <- read_xpt(
  "LLCP2021.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1",
                 "_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2021)


# 2020

BRFSS_2020 <- read_xpt(
  "LLCP2020.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1",
                 "_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2020)


# 2019

BRFSS_2019 <- read_xpt(
  "LLCP2019.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1", "_PA150R3",
                 "_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2019)


# 2018

BRFSS_2018 <- read_xpt(
  "LLCP2018.XPT",
  col_select = c("_STATE","SEX1","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1",
                 "_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
                 "_RFSMOK3","_LLCPWT","_STSTR","_PSU")
) %>% mutate(year = 2018)
```



```{r}

BRFSS <- bind_rows(BRFSS_2018, BRFSS_2019, BRFSS_2020, BRFSS_2021, 
                   BRFSS_2022, BRFSS_2023, BRFSS_2024)

# Remove yearly files after stacking to reduce memory use
rm(BRFSS_2018, BRFSS_2019, BRFSS_2020, BRFSS_2021,
  BRFSS_2022, BRFSS_2023, BRFSS_2024)
```



```{r}

# Harmonize analysis variables across BRFSS survey years.
# This step creates consistent variables for:
# - state
# - sex
# - age and age group
# - race/ethnicity
# - education
# - insurance
# - personal health care provider
# - marital status
# - smoking status
# - physical activity
# - aerobic guideline adherence (2019 and 2023 only)
# - cancer history



BRFSS <- BRFSS %>%
  dplyr::mutate(
    
    # State
    State = dplyr::case_when(
      `_STATE` == 1  ~ "Alabama",
      `_STATE` == 2  ~ "Alaska",
      `_STATE` == 4  ~ "Arizona",
      `_STATE` == 5  ~ "Arkansas",
      `_STATE` == 6  ~ "California",
      `_STATE` == 8  ~ "Colorado",
      `_STATE` == 9  ~ "Connecticut",
      `_STATE` == 10 ~ "Delaware",
      `_STATE` == 11 ~ "District of Columbia",
      `_STATE` == 12 ~ "Florida",
      `_STATE` == 13 ~ "Georgia",
      `_STATE` == 15 ~ "Hawaii",
      `_STATE` == 16 ~ "Idaho",
      `_STATE` == 17 ~ "Illinois",
      `_STATE` == 18 ~ "Indiana",
      `_STATE` == 19 ~ "Iowa",
      `_STATE` == 20 ~ "Kansas",
      `_STATE` == 21 ~ "Kentucky",
      `_STATE` == 22 ~ "Louisiana",
      `_STATE` == 23 ~ "Maine",
      `_STATE` == 24 ~ "Maryland",
      `_STATE` == 25 ~ "Massachusetts",
      `_STATE` == 26 ~ "Michigan",
      `_STATE` == 27 ~ "Minnesota",
      `_STATE` == 28 ~ "Mississippi",
      `_STATE` == 29 ~ "Missouri",
      `_STATE` == 30 ~ "Montana",
      `_STATE` == 31 ~ "Nebraska",
      `_STATE` == 32 ~ "Nevada",
      `_STATE` == 33 ~ "New Hampshire",
      `_STATE` == 34 ~ "New Jersey",
      `_STATE` == 35 ~ "New Mexico",
      `_STATE` == 36 ~ "New York",
      `_STATE` == 37 ~ "North Carolina",
      `_STATE` == 38 ~ "North Dakota",
      `_STATE` == 39 ~ "Ohio",
      `_STATE` == 40 ~ "Oklahoma",
      `_STATE` == 41 ~ "Oregon",
      `_STATE` == 42 ~ "Pennsylvania",
      `_STATE` == 44 ~ "Rhode Island",
      `_STATE` == 45 ~ "South Carolina",
      `_STATE` == 46 ~ "South Dakota",
      `_STATE` == 47 ~ "Tennessee",
      `_STATE` == 48 ~ "Texas",
      `_STATE` == 49 ~ "Utah",
      `_STATE` == 50 ~ "Vermont",
      `_STATE` == 51 ~ "Virginia",
      `_STATE` == 53 ~ "Washington",
      `_STATE` == 54 ~ "West Virginia",
      `_STATE` == 55 ~ "Wisconsin",
      `_STATE` == 56 ~ "Wyoming",
      `_STATE` == 60 ~ "American Samoa",
      `_STATE` == 66 ~ "Guam",
      `_STATE` == 69 ~ "Northern Mariana Islands",
      `_STATE` == 72 ~ "Puerto Rico",
      `_STATE` == 78 ~ "U.S. Virgin Islands",
      TRUE ~ NA_character_
    ),

    
    # Sex
    sex = dplyr::case_when(
      year == 2018 & SEX1   == 1 ~ "Male",
      year == 2018 & SEX1   == 2 ~ "Female",
      year >= 2019 & SEXVAR == 1 ~ "Male",
      year >= 2019 & SEXVAR == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    sex = factor(sex, levels = c("Male", "Female")),

    
    # Age
    age = `_AGE80`,
    age_categorical = cut(
      `_AGE80`,
      breaks = c(-Inf, 49, 59, 69, 79, Inf),
      labels = c("<50", "50–59", "60–69", "70–79", "80+"),
      right = TRUE
    ),
    age_categorical = factor(
      age_categorical,
      levels = c("<50", "50–59", "60–69", "70–79", "80+"),
      ordered = TRUE
    ),

    
    # Race / Ethnicity
    race_source = dplyr::case_when(
      !is.na(`_RACE`)    ~ "RACE",
      !is.na(`_RACEPRV`) ~ "RACEPRV",
      !is.na(`_RACEPR1`) ~ "RACEPR1",
      TRUE ~ NA_character_
    ),
    race_code = dplyr::case_when(
      race_source == "RACE"    ~ as.character(`_RACE`),
      race_source == "RACEPRV" ~ as.character(`_RACEPRV`),
      race_source == "RACEPR1" ~ as.character(`_RACEPR1`),
      TRUE ~ NA_character_
    ),
    race_ethnicity = dplyr::case_when(
      race_code == "1" ~ "Non-Hispanic White",
      race_code == "2" ~ "Non-Hispanic Black",
      race_code == "3" ~ "Non-Hispanic American Indian/Alaska Native",
      race_code == "4" ~ "Non-Hispanic Asian",
      race_code == "5" ~ "Non-Hispanic Asian Pacific Islander",

      # Multiracial Logic:
      race_source %in% c("RACE", "RACEPRV") & race_code %in% c("6", "7") ~ "Non-Hispanic Other/Multiracial",
      race_source == "RACEPR1" & race_code == "6"                        ~ "Non-Hispanic Other/Multiracial",

      # Hispanic Logic:
      race_source %in% c("RACE", "RACEPRV") & race_code == "8" ~ "Hispanic",
      race_source == "RACEPR1" & race_code == "7"              ~ "Hispanic",

      TRUE ~ NA_character_
    ),
    race_ethnicity = factor(
      race_ethnicity,
      levels = c(
        "Non-Hispanic White",
        "Non-Hispanic Black",
        "Non-Hispanic American Indian/Alaska Native",
        "Non-Hispanic Asian",
        "Non-Hispanic Asian Pacific Islander",
        "Non-Hispanic Other/Multiracial",
        "Hispanic"
      )
    ),

    
    # Education
    education = dplyr::case_when(
      EDUCA %in% c(1, 2, 3) ~ "Less than high school",
      EDUCA == 4            ~ "High school or GED",
      EDUCA == 5            ~ "Some college",
      EDUCA == 6            ~ "College or more",
      TRUE ~ NA_character_
    ),
    education = factor(
      education,
      levels = c("Less than high school", "High school or GED", "Some college", "College or more")
    ),

    
    # Insurance
    insurance_raw = dplyr::coalesce(`_HLTHPL2`, `_HLTHPL1`, `_HLTHPLN`, HLTHPLN1),
    insurance = dplyr::case_when(
      insurance_raw == 1 ~ "Insured",
      insurance_raw == 2 ~ "No insurance",
      TRUE ~ NA_character_
    ),
    insurance = factor(insurance, levels = c("Insured", "No insurance")),

    
    # Personal Health Care Provider
    PHC_provider = dplyr::case_when(
      year %in% 2018:2020 & PERSDOC2 %in% c(1, 2) ~ "Yes",
      year %in% 2018:2020 & PERSDOC2 == 3         ~ "No",
      year %in% 2021:2024 & PERSDOC3 %in% c(1, 2) ~ "Yes",
      year %in% 2021:2024 & PERSDOC3 == 3         ~ "No",
      TRUE ~ NA_character_
    ),
    PHC_provider = factor(PHC_provider, levels = c("Yes", "No")),

   
    # Marital status
    marital_status = dplyr::case_when(
      MARITAL == 1              ~ "Married",
      MARITAL %in% c(2, 3, 4, 5, 6) ~ "Unmarried",
      TRUE ~ NA_character_
    ),
    marital_status = factor(marital_status, levels = c("Married", "Unmarried")),

    
    # Smoking status
    smoking_status = dplyr::case_when(
      `_RFSMOK3` == 2 ~ "Current smoker",
      `_RFSMOK3` == 1 ~ "Not current smoker",
      TRUE ~ NA_character_
    ),
    smoking_status = factor(smoking_status, levels = c("Current smoker", "Not current smoker")),

   
     # Physical activity
    physical_activity = dplyr::case_when(
      EXERANY2 %in% c(7, 9) ~ NA_real_,
      EXERANY2 == 1 ~ 1,
      EXERANY2 == 2 ~ 0,
      TRUE ~ NA_real_
    ),

    
    # Aerobic physical activity guideline adherence (2019 & 2023 Only)
    aerobic_guideline = dplyr::case_when(
      year == 2019 & `_PA150R3` == 1           ~ "Met guidelines",
      year == 2019 & `_PA150R3` %in% c(2, 3)   ~ "Did not meet",
      year == 2023 & `_PA150R4` == 1           ~ "Met guidelines",
      year == 2023 & `_PA150R4` %in% c(2, 3)   ~ "Did not meet",
      TRUE ~ NA_character_
    ),
    aerobic_guideline = factor(aerobic_guideline, levels = c("Did not meet", "Met guidelines")),

   
    # Cancer history
    # Convert 7 (Don't know) and 9 (Refused) to NA
    
    CHCOCNCR = dplyr::if_else(CHCOCNCR %in% c(7, 9), NA_real_, CHCOCNCR),
    CHCOCNC1 = dplyr::if_else(CHCOCNC1 %in% c(7, 9), NA_real_, CHCOCNC1),

    cancer_history = dplyr::case_when(
      year %in% 2018:2021 & CHCOCNCR == 1 ~ "Yes",
      year %in% 2018:2021 & CHCOCNCR == 2 ~ "No",
      year %in% 2022:2024 & CHCOCNC1 == 1 ~ "Yes",
      year %in% 2022:2024 & CHCOCNC1 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    cancer_history = factor(cancer_history, levels = c("No", "Yes"))
  ) %>%
  
  
  # Rename survey design variables for consistency across analyses
  dplyr::rename(
    PSU    = `_PSU`,
    STRATA = `_STSTR`,
    WEIGHT = `_LLCPWT`
  ) %>%
  # Drop helper columns
  dplyr::select(-race_source, -race_code, -insurance_raw)
```


 Missingness assessment
 Summarize missingness in the outcome, exposure, covariates, and survey design variables.

```{r}

# Summarize missingness in the outcome, exposure, covariates, and survey design variables.
# (Supplementary table 3A)

tibble::tibble(
  Variable = c(
    "Smoking status",
    "Insurance",
    "Marital status",
    "Personal health care provider",
    "Education",
    "Physical activity",
    "Race/Ethnicity",
    "Sex",
    "Cancer history",
    "Age group",
    "Year",
    "WEIGHT",
    "STRATA",
    "PSU"
  ),
  Total_n = nrow(BRFSS),
  N_missing = c(
    sum(is.na(BRFSS$smoking_status)),
    sum(is.na(BRFSS$insurance)),
    sum(is.na(BRFSS$marital_status)),
    sum(is.na(BRFSS$PHC_provider)),
    sum(is.na(BRFSS$education)),
    sum(is.na(BRFSS$physical_activity)),
    sum(is.na(BRFSS$race_ethnicity)),
    sum(is.na(BRFSS$sex)),
    sum(is.na(BRFSS$cancer_history)),
    sum(is.na(BRFSS$age_categorical)),
    sum(is.na(BRFSS$year)),
    sum(is.na(BRFSS$WEIGHT)),
    sum(is.na(BRFSS$STRATA)),
    sum(is.na(BRFSS$PSU))
  )
) %>%
  dplyr::mutate(
    pct_missing = round(100 * N_missing / Total_n, 2)
  ) %>%
  dplyr::arrange(dplyr::desc(pct_missing))
```



```{r}

# Missingness in key variables by survey year

missing_by_year <- BRFSS %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(
    total = dplyr::n(),
    missing_physical_activity = sum(is.na(physical_activity)),
    percent_missing_physical_activity = round(100 * missing_physical_activity / total, 2),
    missing_cancer_history = sum(is.na(cancer_history)),
    percent_missing_cancer_history = round(100 * missing_cancer_history / total, 2),
    .groups = "drop"
  ) %>%
  dplyr::arrange(year)

missing_by_year
```


```{r}

# State-level missingness in key variables for 2019 (diagnostic check)

missing_by_state_2019 <- BRFSS %>%
  dplyr::filter(year == 2019) %>%
  dplyr::group_by(State) %>%
  dplyr::summarise(
    total = dplyr::n(),
    missing_physical_activity = sum(is.na(physical_activity)),
    percent_missing_physical_activity = round(100 * missing_physical_activity / total, 2),
    missing_cancer_history = sum(is.na(cancer_history)),
    percent_missing_cancer_history = round(100 * missing_cancer_history / total, 2),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(percent_missing_physical_activity))
missing_by_state_2019
```



 Create complete-case analytic dataset for the main cancer history analysis.
 This dataset includes the outcome, exposure, covariates, and survey design variables.

```{r}

analytic_main <- BRFSS %>%
  tidyr::drop_na(
    physical_activity, cancer_history, age_categorical, sex, race_ethnicity,
    education, insurance, PHC_provider, marital_status, smoking_status, 
    year, WEIGHT, STRATA, PSU
  ) %>%
  mutate(
    year = factor(year, levels = 2018:2024),
    
    # Create unique strata IDs
    
    # BRFSS strata and PSU codes are assigned independently each year, so the same
    # numeric code can appear across years but refer to different design elements.
    # Interacting year with STRATA and PSU creates unique identifiers that prevent
    # the survey package from incorrectly treating codes from different years as the
    # same stratum or PSU in the pooled design.
    
    unique_strata = interaction(year, STRATA),
    unique_psu    = interaction(year, PSU)
  )

options(survey.lonely.psu = "adjust")

# Create pooled survey design object for the main analysis.

design_main <- svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata, 
  weights = ~WEIGHT,
  data    = analytic_main,
  nest    = TRUE
)

```


# Survey-weighted prevalence and trends

```{r}

# Estimate annual survey-weighted prevalence of physical activity by cancer history.

prev_trend_cc <- svyby(
  ~physical_activity,
  ~year + cancer_history,
  design_main,
  svymean,
  na.rm   = TRUE,
  vartype = c("ci")
) %>%
  dplyr::mutate(                              
    prevalence_pct = round(physical_activity * 100, 1),
    ci_low_pct     = round(ci_l * 100, 1),
    ci_high_pct    = round(ci_u * 100, 1)
  )

prev_trend_cc
```


```{r}

# Add numeric survey year for plotting and trend models.

prev_trend_cc <- prev_trend_cc %>%
  dplyr::mutate(year_num = as.numeric(as.character(year)))
```


```{r}

# Create formatted table of annual weighted prevalence by cancer history.
# Output included in supplementary table 4.

table_prev_cc <- prev_trend_cc %>%
  mutate(
    pa_ci = paste0(
      round(prevalence_pct, 1), "% (",
      round(ci_low_pct, 1), "–",
      round(ci_high_pct, 1), "%)"
    )
  ) %>%
  select(year, cancer_history, pa_ci) %>%
  pivot_wider(                                                                           
    names_from  = year,
    values_from = pa_ci
  ) %>%
  arrange(match(cancer_history, c("Yes", "No"))) %>%
  rename(`Cancer history` = cancer_history)

table_prev_cc %>%
  kable(
    caption = "Weighted prevalence of physical activity (%) by cancer history and survey year, BRFSS 2018–2024"
  )
table_prev_cc
```


 Estimate annual percentage change (APC) in physical activity prevalence
 separately for adults with and without a cancer history
 
```{r}

# Output included in table 2.

trend_yes <- prev_trend_cc %>% dplyr::filter(cancer_history == "Yes")
trend_no  <- prev_trend_cc %>% dplyr::filter(cancer_history == "No")


lm_log_yes <- lm(log(physical_activity) ~ year_num, data = trend_yes)
lm_log_no  <- lm(log(physical_activity) ~ year_num, data = trend_no)

# Extract APC with 95% CIs
apc_results <- dplyr::bind_rows(
  broom::tidy(lm_log_yes, conf.int = TRUE) %>%
    dplyr::filter(term == "year_num") %>%
    dplyr::mutate(group = "Cancer history: Yes"),
  broom::tidy(lm_log_no, conf.int = TRUE) %>%
    dplyr::filter(term == "year_num") %>%
    dplyr::mutate(group = "Cancer history: No")
) %>%
  dplyr::transmute(
    group,
    APC_pct     = round((exp(estimate) - 1) * 100, 2),
    ci_low_pct  = round((exp(conf.low) - 1) * 100, 2),
    ci_high_pct = round((exp(conf.high) - 1) * 100, 2),
    p_value     = p.value
  )

apc_results
```


 Estimate the adjusted prevalence difference in physical activity
 between adults with and without a cancer history
 
```{r}
# aPD model
lpm_overall <- survey::svyglm(
  physical_activity ~ cancer_history +
    age_categorical + sex + race_ethnicity +
    education + insurance + PHC_provider +
    marital_status + smoking_status + year,
  design = design_main,
  family = gaussian(link = "identity")
)

# Extract aPD
rd_overall <- broom::tidy(lpm_overall, conf.int = TRUE) %>%
  dplyr::filter(term == "cancer_historyYes") %>%
  dplyr::transmute(
    comparison = "Cancer history: Yes vs No",
    aPD_pp     = round(estimate * 100, 1),
    ci_low_pp  = round(conf.low * 100, 1),
    ci_high_pp = round(conf.high * 100, 1),
    p_value    = p.value
  )
rd_overall
```


 Create the main figure showing annual weighted prevalence of 
 physical activity by cancer history.

```{r}

# Output included on manuscript as figure 1.

prev_trend_cc <- prev_trend_cc %>%
  dplyr::mutate(
    year_num = as.numeric(as.character(year)),
    cancer_history = factor(cancer_history, levels = c("Yes", "No"))
  )

gg_pa_trend <- ggplot(
  prev_trend_cc,
  aes(
    x = year_num,
    y = prevalence_pct,
    group = cancer_history,
    linetype = cancer_history,
    shape = cancer_history
  )
) +
  geom_errorbar(
    aes(ymin = ci_low_pct, ymax = ci_high_pct),
    width = 0.15,
    linewidth = 0.45
  ) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5, fill = "white", stroke = 0.8) +
  scale_x_continuous(breaks = 2018:2024) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10),
    expand = expansion(mult = c(0, 0.02)),
    name = "Weighted prevalence of PA (%)"
  ) +
  scale_linetype_manual(
    name = "Cancer history",
    values = c("No" = "22", "Yes" = "solid")
  ) +
  scale_shape_manual(
    name = "Cancer history",
    values = c("No" = 17, "Yes" = 16)
  ) +
  labs(x = "Survey year") +
  theme_classic(base_size = 12) +
  theme(
    axis.text = element_text(color = "black"),
    legend.text = element_text(size = 11),
    legend.key.width = unit(2, "lines"),
    legend.position = "top"
  )

gg_pa_trend
```


```{r}
ggsave(
  filename = "Fig1_PA_trend_cancer_history.png",
  plot     = gg_pa_trend,
  width    = 6.5,   
  height   = 4.5,   
  units    = "in",
  dpi      = 1200    
)
```




```{r}

# Create Table 1 showing unweighted participant characteristics by cancer history.

# Recode physical activity to a labeled factor for the table
analytic_main_tab <- analytic_main %>%
  dplyr::mutate(
    physical_activity = factor(
      physical_activity,
      levels = c(1, 0),
      labels = c("Yes", "No")
    )
  )

# Table 1: unweighted counts and column percentages by cancer history
table1_main <- analytic_main_tab %>%
  tbl_summary(
    by        = cancer_history,
    include   = c(physical_activity,age_categorical,sex,race_ethnicity,education,insurance,marital_status,smoking_status,PHC_provider),
    statistic = all_categorical() ~ "{n} ({p}%)",
    percent   = "column",
    missing   = "ifany",
    digits    = all_categorical() ~ 1
  ) %>%
  add_overall() %>%                 # Overall column
  add_n(location = "header") %>%    # N in each column header
  modify_header(label ~ "**Characteristic**") %>%
  modify_caption( "**Table 1.  Unweighted sociodemographic and behavioral characteristics of adults in the analytic
sample by self-reported cancer history, BRFSS 2018–2024**")

table1_main
```





 Aerobic physical activity guideline subanalysis
 Assess missingness and estimate weighted prevalence of meeting and not meeting
 the aerobic guideline by cancer history in 2019 and 2023.

```{r}

# Summarize complete cases for the aerobic-guideline subanalysis.
BRFSS %>%
  dplyr::filter(year %in% c(2019, 2023)) %>%
  summarise(
    total_n = dplyr::n(),
    missing_any = sum(
      is.na(aerobic_guideline) |
      is.na(cancer_history) |
      is.na(WEIGHT) |
      is.na(STRATA) |
      is.na(PSU)
    ),
    complete_cases = sum(
      !is.na(aerobic_guideline) &
      !is.na(cancer_history) &
      !is.na(WEIGHT) &
      !is.na(STRATA) &
      !is.na(PSU)
    )
  )
```


```{r}

# Summarize missingness in key variables by survey year.
BRFSS %>%
  dplyr::filter(year %in% c(2019, 2023)) %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(
    total_n = dplyr::n(),
    missing_aerobic_guideline = sum(is.na(aerobic_guideline)),
    missing_cancer_history = sum(is.na(cancer_history)),
    missing_WEIGHT = sum(is.na(WEIGHT)),
    missing_STRATA = sum(is.na(STRATA)),
    missing_PSU = sum(is.na(PSU)),
    missing_any = sum(
      is.na(aerobic_guideline) |
      is.na(cancer_history) |
      is.na(WEIGHT) |
      is.na(STRATA) |
      is.na(PSU)
    )
  )
```


```{r}

# Create analytic dataset and survey design for the aerobic-guideline subanalysis.
guideline_df <- BRFSS %>%
  dplyr::filter(year %in% c(2019, 2023)) %>%
  tidyr::drop_na(aerobic_guideline, cancer_history, WEIGHT, STRATA, PSU) %>%
  dplyr::mutate(
    unique_strata = interaction(year, STRATA),
    unique_psu    = interaction(year, PSU)
  )

options(survey.lonely.psu = "adjust")
guideline_design <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = guideline_df,
  nest    = TRUE
)

guideline_prev <- survey::svyby(
  ~aerobic_guideline,
  ~year + cancer_history,
  guideline_design,
  survey::svymean,
  vartype = "ci"
) %>%
  tibble::as_tibble()

guideline_prev
```



```{r}

# Create formatted table of aerobic guideline adherence by cancer history.
# Output included as table 3 on the manuscript.


guideline_table_full <- guideline_prev %>%
  dplyr::mutate(
    met_stats = paste0(
      round(`aerobic_guidelineMet guidelines` * 100, 1), "% (",
      round(`ci_l.aerobic_guidelineMet guidelines` * 100, 1), "–",
      round(`ci_u.aerobic_guidelineMet guidelines` * 100, 1), "%)"
    ),
    not_met_stats = paste0(
      round(`aerobic_guidelineDid not meet` * 100, 1), "% (",
      round(`ci_l.aerobic_guidelineDid not meet` * 100, 1), "–",
      round(`ci_u.aerobic_guidelineDid not meet` * 100, 1), "%)"
    )
  ) %>%
  dplyr::select(year, cancer_history, met_stats, not_met_stats) %>%
  dplyr::arrange(year, cancer_history) %>%
  dplyr::rename(
    `Cancer history` = cancer_history,
    `Met guideline (≥150 min/week)` = met_stats,
    `Did not meet guideline (<150 min/week)` = not_met_stats
  )

knitr::kable(
  guideline_table_full,
  caption = "Weighted prevalence of meeting and not meeting the aerobic guideline by cancer history in 2019 and 2023"
)

guideline_table_full
```


 Supplemental data checks and flow-diagram counts

```{r}

# Discordant cancer type reports by sex
BRFSS %>%
  dplyr::summarise(
    `Male respondents reporting breast cancer` = sum(
      cancer_history == "Yes" & sex == "Male" &
        ((year %in% 2018:2021 & CNCRTYP1 == 1) |
         (year %in% 2022:2024 & CNCRTYP2 == 5)),
      na.rm = TRUE
    ),
    `Female respondents reporting prostate cancer` = sum(
      cancer_history == "Yes" & sex == "Female" &
        ((year %in% 2018:2021 & CNCRTYP1 == 19) |
         (year %in% 2022:2024 & CNCRTYP2 == 20)),
      na.rm = TRUE
    )
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "check",
    values_to = "discordant_n"
  ) %>%
  knitr::kable(caption = "Discordant sex-specific cancer type reports")
```



 Sample derivation for the primary dataset 

```{r}

# Summarize sequential exclusions used to create the complete-case analytic dataset.
# Output used in supplementary figure 1.

tibble::tibble(
  step = c(
    "Adults in BRFSS 2018–2024",
    "Excluded: missing physical activity",
    "Excluded: missing cancer history",
    "Excluded: missing covariates or survey design variables",
    "Final analytic dataset"
  ),
  n = c(
    nrow(BRFSS),
    nrow(BRFSS) - nrow(BRFSS %>% dplyr::filter(!is.na(physical_activity))),
    nrow(BRFSS %>% dplyr::filter(!is.na(physical_activity))) -
      nrow(BRFSS %>% dplyr::filter(!is.na(physical_activity), !is.na(cancer_history))),
    nrow(BRFSS %>% dplyr::filter(!is.na(physical_activity), !is.na(cancer_history))) -
      nrow(analytic_main),
    nrow(analytic_main)
  )
) %>%
  knitr::kable(caption = "Sample derivation for the main cancer history analysis")
```

