---
title: "Cancer type-specific analysis of Physical Activity Using BRFSS, 2018–2024"
author: "Biruk Demma"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



This script analyzes BRFSS 2018–2024 data from states that administered the
optional cancer type module to:

 1. Estimate survey-weighted prevalence of physical activity by cancer type.
 2. Evaluate temporal trends in physical activity prevalence using annual
    percentage change (APC).
 3. Estimate adjusted prevalence differences in physical activity by cancer type
    compared with adults without a cancer history.


```{r}

#load packages
pacman::p_load(haven, dplyr, tidyr, tibble, survey, broom, ggplot2, scales, knitr, stringr)
```


BRFSS data Import

 BRFSS data are distributed in multiple release files:
 - The main file (LLCP2024.XPT) includes all states in the core component.
 - The V1, V2, and V3 files include separately released state data with different
   final weight variables.

```{r}

# 2024 DATA

BRFSS_t_24_main <- read_xpt("LLCP2024.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPL2","MARITAL","PERSDOC3",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU" )) %>%
  filter(`_STATE` %in% c(9,10,13,18,25,26,29,34,37,46,49,78,51)) %>%
  mutate(year = 2024)

BRFSS_t_24_v1 <- read_xpt("LLCP24V1.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2", "_AGE80","_AGE_G","EDUCA","_HLTHPL2","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV1","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(4,20)) %>%
  mutate(year = 2024)

BRFSS_t_24_v2 <- read_xpt("LLCP24V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPL2","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(31,36)) %>%
  mutate(year = 2024)

BRFSS_t_2024 <- bind_rows(BRFSS_t_24_main, BRFSS_t_24_v1, BRFSS_t_24_v2)

rm(BRFSS_t_24_main, BRFSS_t_24_v1, BRFSS_t_24_v2)
```


```{r}

# 2023 DATA

BRFSS_t_23_main <- read_xpt("LLCP2023.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2","_PA150R4","_AGE80","_AGE_G","EDUCA","_HLTHPL1","MARITAL",
                 "PERSDOC3", "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(9,13,17,18,26,27,28,33,34,47,78,51,54,55)) %>%
  mutate(year = 2023)

BRFSS_t_23_v1 <- read_xpt("LLCP23V1.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2","_PA150R4","_AGE80","_AGE_G","EDUCA","_HLTHPL1","MARITAL",
                 "PERSDOC3", "_RFSMOK3","_LCPWTV1","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(36,40)) %>%
  mutate(year = 2023)

BRFSS_t_23_v2 <- read_xpt("LLCP23V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNC1","CNCRTYP2","_PA150R4","_AGE80","_AGE_G","EDUCA","_HLTHPL1","MARITAL",
                 "PERSDOC3","_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(20,24)) %>%
  mutate(year = 2023)

BRFSS_t_2023 <- bind_rows(BRFSS_t_23_main, BRFSS_t_23_v1, BRFSS_t_23_v2)

rm(BRFSS_t_23_main, BRFSS_t_23_v1, BRFSS_t_23_v2)
```


```{r}

# 2022 DATA

BRFSS_t_22_main <- read_xpt("LLCP2022.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPR1","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(5,8,9,10,15,16,17,18,26,29,31,32,34,35,37,41,72,44,46,49,50,51,55)) %>%
  mutate(year = 2022)

BRFSS_t_22_v1 <- read_xpt("LLCP22V1.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPR1","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV1","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(19,25,39,40)) %>%
  mutate(year = 2022)

BRFSS_t_22_v2 <- read_xpt("LLCP22V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPR1","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(4,20)) %>%
  mutate(year = 2022)

BRFSS_t_22_v3 <- read_xpt("LLCP22V3.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPR1","CHCOCNC1","CNCRTYP2","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV3","_STSTR","_PSU")) %>%
  filter(`_STATE` == 24) %>%
  mutate(year = 2022)

BRFSS_t_2022 <- bind_rows(BRFSS_t_22_main, BRFSS_t_22_v1, BRFSS_t_22_v2, BRFSS_t_22_v3)

rm(BRFSS_t_22_main, BRFSS_t_22_v1, BRFSS_t_22_v2, BRFSS_t_22_v3)
```


```{r}

# 2021 DATA

BRFSS_t_21_main <- read_xpt(
  "LLCP2021.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(5,10,15,26,30,34,35,72,49,55)) %>%
  mutate(year = 2021)

BRFSS_t_21_v1 <- read_xpt("LLCP21V1.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV1","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(31,40)) %>%
  mutate(year = 2021)

BRFSS_t_21_v2 <- read_xpt("LLCP21V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` == 20) %>%
  mutate(year = 2021)

BRFSS_t_21_v3 <- read_xpt("LLCP21V3.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","_HLTHPLN","MARITAL","PERSDOC3",
    "_RFSMOK3","_LCPWTV3","_STSTR","_PSU")) %>%
  filter(`_STATE` == 24) %>%
  mutate(year = 2021)

BRFSS_t_2021 <- bind_rows(BRFSS_t_21_main, BRFSS_t_21_v1, BRFSS_t_21_v2, BRFSS_t_21_v3)

rm(BRFSS_t_21_main, BRFSS_t_21_v1, BRFSS_t_21_v2, BRFSS_t_21_v3)
```


```{r}

# 2020 DATA

BRFSS_t_20_main <- read_xpt("LLCP2020.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(4,9,10,13,66,15,18,22,25,26,28,29,30,34,35,37,44,46,49,50,51,55)) %>%
  mutate(year = 2020)

BRFSS_t_20_v1 <- read_xpt("LLCP20V1.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LCPWTV1","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(20,23,31,36,39,40)) %>%
  mutate(year = 2020)

BRFSS_t_20_v2 <- read_xpt("LLCP20V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` == 24) %>%
  mutate(year = 2020)

BRFSS_t_20_v3 <- read_xpt("LLCP20V3.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACEPRV","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LCPWTV3","_STSTR","_PSU")) %>%
  filter(`_STATE` == 36) %>%
  mutate(year = 2020)

BRFSS_t_2020 <- bind_rows(BRFSS_t_20_main, BRFSS_t_20_v1, BRFSS_t_20_v2, BRFSS_t_20_v3)

rm(BRFSS_t_20_main, BRFSS_t_20_v1, BRFSS_t_20_v2, BRFSS_t_20_v3)
```


```{r}

# 2019 DATA

BRFSS_t_19_main <- read_xpt("LLCP2019.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1","_PA150R3","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` == 56) %>%
  mutate(year = 2019)

BRFSS_t_19_v2 <- read_xpt("LLCP19V2.XPT",
  col_select = c("_STATE","SEXVAR","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1","_PA150R3","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(24,25)) %>%
  mutate(year = 2019)

BRFSS_t_2019 <- bind_rows(BRFSS_t_19_main, BRFSS_t_19_v2)

rm(BRFSS_t_19_main, BRFSS_t_19_v2)
```


```{r}

# 2018 DATA

BRFSS_t_18_main <- read_xpt("LLCP2018.XPT",
  col_select = c("_STATE","SEX1","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LLCPWT","_STSTR","_PSU")) %>%
  filter(`_STATE` %in% c(10,18,26,29,34,46)) %>%
  mutate(year = 2018)

BRFSS_t_18_v2 <- read_xpt("LLCP18V2.XPT",
  col_select = c("_STATE","SEX1","EXERANY2","_RACE","CHCOCNCR","CNCRTYP1","_AGE80","_AGE_G","EDUCA","HLTHPLN1","MARITAL","PERSDOC2",
    "_RFSMOK3","_LCPWTV2","_STSTR","_PSU")) %>%
  filter(`_STATE` == 20) %>%
  mutate(year = 2018)

BRFSS_t_2018 <- bind_rows(BRFSS_t_18_main, BRFSS_t_18_v2)

rm(BRFSS_t_18_main, BRFSS_t_18_v2)
```


```{r}

# Stacked annual datasets

BRFSS_t <- bind_rows( BRFSS_t_2018, BRFSS_t_2019, BRFSS_t_2020, BRFSS_t_2021, BRFSS_t_2022, BRFSS_t_2023, BRFSS_t_2024)

rm(BRFSS_t_2018, BRFSS_t_2019, BRFSS_t_2020, BRFSS_t_2021, BRFSS_t_2022, BRFSS_t_2023, BRFSS_t_2024)
```



# Weight Consolidation

```{r}

# Identify the BRFSS release file for each record and create a single final weight variable
# that harmonizes the file-specific weight variables across main, V1, V2, and V3 releases.

BRFSS_t <- BRFSS_t %>%
  dplyr::mutate(
    version = dplyr::case_when(
      !is.na(`_LLCPWT`)  ~ "main",
      !is.na(`_LCPWTV1`) ~ "v1",
      !is.na(`_LCPWTV2`) ~ "v2",
      !is.na(`_LCPWTV3`) ~ "v3",
      TRUE ~ NA_character_),
    
    WEIGHT = dplyr::coalesce(`_LLCPWT`, `_LCPWTV1`, `_LCPWTV2`, `_LCPWTV3`)) %>%
    dplyr::mutate(year = as.integer(year),
               version = factor(version, levels = c("main","v1","v2","v3")))
```


```{r}

# Diagnostic check: 
# Identify any state-year combinations that appear in more than one BRFSS_t release version (main/v1/v2/v3). 
# If any are found, those state-years require weight adjustments.

BRFSS_t %>%
  dplyr::filter(!is.na(version)) %>%
  dplyr::distinct(year, `_STATE`, version) %>%
  dplyr::count(year, `_STATE`, name = "n_versions") %>%
  dplyr::filter(n_versions > 1) %>%
  dplyr::arrange(year, `_STATE`)
```


```{r}

# For State 36 (NY) in 2020 only: compute N’s and rescale weights for v1 vs v3 proportionally.

NYn2020 <- BRFSS_t %>%
  dplyr::filter(year == 2020, `_STATE` == 36, version %in% c("v1","v3")) %>%
  dplyr::count(version, name = "n") %>%
  tidyr::pivot_wider(names_from = version, values_from = n, values_fill = 0)

N1 <- as.numeric(NYn2020$v1)
N3 <- as.numeric(NYn2020$v3)
P1 <- N1 / (N1 + N3)
P3 <- N3 / (N1 + N3)

BRFSS_t <- BRFSS_t %>%
  dplyr::mutate( WEIGHT = dplyr::case_when(
      year == 2020 & `_STATE` == 36 & version == "v1" ~ WEIGHT * P1,
      year == 2020 & `_STATE` == 36 & version == "v3" ~ WEIGHT * P3,
      TRUE ~ WEIGHT))
NYn2020
P1
P3
```

```{r}

# Harmonization

BRFSS_t <- BRFSS_t %>%
  dplyr::mutate(
    year_num = as.integer(year),

    
    # STATE
    State = dplyr::case_when(
      `_STATE` == 1  ~ "Alabama", `_STATE` == 2  ~ "Alaska", `_STATE` == 4  ~ "Arizona",
      `_STATE` == 5  ~ "Arkansas", `_STATE` == 6  ~ "California", `_STATE` == 8  ~ "Colorado",
      `_STATE` == 9  ~ "Connecticut", `_STATE` == 10 ~ "Delaware", `_STATE` == 11 ~ "District of Columbia",
      `_STATE` == 12 ~ "Florida", `_STATE` == 13 ~ "Georgia", `_STATE` == 15 ~ "Hawaii",
      `_STATE` == 16 ~ "Idaho", `_STATE` == 17 ~ "Illinois", `_STATE` == 18 ~ "Indiana",
      `_STATE` == 19 ~ "Iowa", `_STATE` == 20 ~ "Kansas", `_STATE` == 21 ~ "Kentucky",
      `_STATE` == 22 ~ "Louisiana", `_STATE` == 23 ~ "Maine", `_STATE` == 24 ~ "Maryland",
      `_STATE` == 25 ~ "Massachusetts", `_STATE` == 26 ~ "Michigan", `_STATE` == 27 ~ "Minnesota",
      `_STATE` == 28 ~ "Mississippi", `_STATE` == 29 ~ "Missouri", `_STATE` == 30 ~ "Montana",
      `_STATE` == 31 ~ "Nebraska", `_STATE` == 32 ~ "Nevada", `_STATE` == 33 ~ "New Hampshire",
      `_STATE` == 34 ~ "New Jersey", `_STATE` == 35 ~ "New Mexico", `_STATE` == 36 ~ "New York",
      `_STATE` == 37 ~ "North Carolina", `_STATE` == 38 ~ "North Dakota", `_STATE` == 39 ~ "Ohio",
      `_STATE` == 40 ~ "Oklahoma", `_STATE` == 41 ~ "Oregon", `_STATE` == 42 ~ "Pennsylvania",
      `_STATE` == 44 ~ "Rhode Island", `_STATE` == 45 ~ "South Carolina", `_STATE` == 46 ~ "South Dakota",
      `_STATE` == 47 ~ "Tennessee", `_STATE` == 48 ~ "Texas", `_STATE` == 49 ~ "Utah",
      `_STATE` == 50 ~ "Vermont", `_STATE` == 51 ~ "Virginia", `_STATE` == 53 ~ "Washington",
      `_STATE` == 54 ~ "West Virginia", `_STATE` == 55 ~ "Wisconsin", `_STATE` == 56 ~ "Wyoming",
      `_STATE` == 60 ~ "American Samoa", `_STATE` == 66 ~ "Guam", `_STATE` == 69 ~ "Northern Mariana Islands",
      `_STATE` == 72 ~ "Puerto Rico", `_STATE` == 78 ~ "U.S. Virgin Islands",
      TRUE ~ NA_character_
    ),
    State = factor(State),

    
    # Sex
    sex = dplyr::case_when(
      year_num == 2018 & SEX1   == 1 ~ "Male",
      year_num == 2018 & SEX1   == 2 ~ "Female",
      year_num >= 2019 & SEXVAR == 1 ~ "Male",
      year_num >= 2019 & SEXVAR == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    sex = factor(sex, levels = c("Male", "Female")),

    
    # Age
    age = `_AGE80`,
    age_categorical = cut(
      `_AGE80`,
      breaks = c(-Inf, 49, 59, 69, 79, Inf),
      labels = c("<50", "50–59", "60–69", "70–79", "80+"),
      right  = TRUE
    ),
    age_categorical = factor(
      age_categorical,
      levels  = c("<50", "50–59", "60–69", "70–79", "80+"),
      ordered = TRUE
    ),

    
    # Race / Ethnicity
    race_source = dplyr::case_when(
      !is.na(`_RACE`)    ~ "RACE",
      !is.na(`_RACEPRV`) ~ "RACEPRV",
      !is.na(`_RACEPR1`) ~ "RACEPR1",
      TRUE ~ NA_character_
    ),
    race_code = dplyr::case_when(
      race_source == "RACE"    ~ as.character(`_RACE`),
      race_source == "RACEPRV" ~ as.character(`_RACEPRV`),
      race_source == "RACEPR1" ~ as.character(`_RACEPR1`),
      TRUE ~ NA_character_
    ),
    race_ethnicity = dplyr::case_when(
      race_code == "1" ~ "Non-Hispanic White",
      race_code == "2" ~ "Non-Hispanic Black",
      race_code == "3" ~ "Non-Hispanic American Indian/Alaska Native",
      race_code == "4" ~ "Non-Hispanic Asian",
      race_code == "5" ~ "Non-Hispanic Asian Pacific Islander",
      race_source %in% c("RACE", "RACEPRV") & race_code %in% c("6", "7") ~ "Non-Hispanic Other/Multiracial",
      race_source == "RACEPR1" & race_code == "6"                       ~ "Non-Hispanic Other/Multiracial",
      race_source %in% c("RACE", "RACEPRV") & race_code == "8"          ~ "Hispanic",
      race_source == "RACEPR1" & race_code == "7"                       ~ "Hispanic",
      TRUE ~ NA_character_
    ),
    race_ethnicity = factor(
      race_ethnicity,
      levels = c(
        "Non-Hispanic White", "Non-Hispanic Black",
        "Non-Hispanic American Indian/Alaska Native",
        "Non-Hispanic Asian", "Non-Hispanic Asian Pacific Islander",
        "Non-Hispanic Other/Multiracial", "Hispanic"
      )
    ),

    
    # Education
    education = dplyr::case_when(
      EDUCA %in% c(1, 2, 3) ~ "< High school",
      EDUCA == 4            ~ "High school or GED",
      EDUCA == 5            ~ "Some college",
      EDUCA == 6            ~ "College or more",
      TRUE ~ NA_character_
    ),
    education = factor(
      education,
      levels = c("< High school", "High school or GED", "Some college", "College or more")
    ),

    
    # Insurance
    insurance_raw = dplyr::coalesce(`_HLTHPL2`, `_HLTHPL1`, `_HLTHPLN`, HLTHPLN1),
    insurance = dplyr::case_when(
      insurance_raw == 1 ~ "Insured",
      insurance_raw == 2 ~ "No insurance",
      TRUE ~ NA_character_
    ),
    insurance = factor(insurance, levels = c("Insured", "No insurance")),

    
    # Personal Health Care Provider
    PHC_provider = dplyr::case_when(
      year_num %in% 2018:2020 & PERSDOC2 %in% c(1, 2) ~ "Yes",
      year_num %in% 2018:2020 & PERSDOC2 == 3         ~ "No",
      year_num %in% 2021:2024 & PERSDOC3 %in% c(1, 2) ~ "Yes",
      year_num %in% 2021:2024 & PERSDOC3 == 3         ~ "No",
      TRUE ~ NA_character_
    ),
    PHC_provider = factor(PHC_provider, levels = c("Yes", "No")),

    
    # Marital Status 
    marital_status = dplyr::case_when(
      MARITAL == 1              ~ "Married",
      MARITAL %in% c(2, 3, 4, 5, 6) ~ "Unmarried",
      TRUE ~ NA_character_
    ),
    marital_status = factor(marital_status, levels = c("Married", "Unmarried")),

    
    # Smoking
    smoking_status = dplyr::case_when(
      `_RFSMOK3` == 2 ~ "Current smoker",
      `_RFSMOK3` == 1 ~ "Not current smoker",
      TRUE ~ NA_character_
    ),
    smoking_status = factor(smoking_status, levels = c("Not current smoker", "Current smoker")),

    
    # Physical Activity 
    physical_activity = dplyr::case_when(
      EXERANY2 %in% c(7, 9) ~ NA_real_,
      EXERANY2 == 1 ~ 1,
      EXERANY2 == 2 ~ 0,
      TRUE ~ NA_real_
    ),

    
    # Cancer History
    CHCOCNCR = dplyr::if_else(CHCOCNCR %in% c(7, 9), NA_real_, as.numeric(CHCOCNCR)),
    CHCOCNC1 = dplyr::if_else(CHCOCNC1 %in% c(7, 9), NA_real_, as.numeric(CHCOCNC1)),
    cancer_history = dplyr::case_when(
      year_num %in% 2018:2021 & CHCOCNCR == 1 ~ "Yes",
      year_num %in% 2018:2021 & CHCOCNCR == 2 ~ "No",
      year_num %in% 2022:2024 & CHCOCNC1 == 1 ~ "Yes",
      year_num %in% 2022:2024 & CHCOCNC1 == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    cancer_history = factor(cancer_history, levels = c("No", "Yes")),

    
    # Cancer type 
    CNCRTYP1 = dplyr::if_else(CNCRTYP1 %in% c(77, 99), NA_real_, as.numeric(CNCRTYP1)),
    CNCRTYP2 = dplyr::if_else(CNCRTYP2 %in% c(77, 99), NA_real_, as.numeric(CNCRTYP2)),
    
    cancer_type = dplyr::case_when(

    # Breast
    cancer_history == "Yes" & year_num %in% 2018:2021 & CNCRTYP1 == 1  & sex == "Female" ~ "Breast cancer",
    cancer_history == "Yes" & year_num %in% 2022:2024 & CNCRTYP2 == 5  & sex == "Female" ~ "Breast cancer",
  
    # Prostate
    cancer_history == "Yes" & year_num %in% 2018:2021 & CNCRTYP1 == 19 & sex == "Male"   ~ "Prostate cancer",
    cancer_history == "Yes" & year_num %in% 2022:2024 & CNCRTYP2 == 20 & sex == "Male"   ~ "Prostate cancer",
  
    # Colon
    cancer_history == "Yes" & year_num %in% 2018:2021 & CNCRTYP1 == 10 ~ "Colon cancer",
    cancer_history == "Yes" & year_num %in% 2022:2024 & CNCRTYP2 == 7  ~ "Colon cancer",
  
    # Other (valid sites excluding the above)
    cancer_history == "Yes" & year_num %in% 2018:2021 & CNCRTYP1 %in% 1:30 & !CNCRTYP1 %in% c(1,10,19) ~ "Other cancer",
    cancer_history == "Yes" & year_num %in% 2022:2024 & CNCRTYP2 %in% 1:30 & !CNCRTYP2 %in% c(5,7,20)  ~ "Other cancer",
  
    TRUE ~ NA_character_
  ),
    
    cancer_type = factor( cancer_type, levels = c("Breast cancer", "Colon cancer", "Prostate cancer", "Other cancer"))
   )%>%
  dplyr::select(-race_source, -race_code, -insurance_raw)
```



```{r}
# Rename survey design variables for consistency across analyses
BRFSS_t <- BRFSS_t %>%
  dplyr::rename(STRATA = `_STSTR`, PSU = `_PSU`)
```



```{r}

# Build analytic dataset for cancer type-specific comparisons
# (adults with a given cancer type vs adults with no cancer history)
analytic_type <- BRFSS_t %>%
  
  # Keep only people with known cancer history (drop missing Yes/No)
  dplyr::filter(!is.na(cancer_history)) %>%
  
  # For respondents without a cancer history, retain all records.
  # For respondents with a cancer history, retain only those with non-missing cancer type.
   dplyr::filter(cancer_history == "No" | (cancer_history == "Yes" & !is.na(cancer_type))) %>%
  
  # Restrict to complete cases for variables used in survey-weighted estimates and models
  tidyr::drop_na(physical_activity,age_categorical,sex, race_ethnicity,education, insurance,
                 PHC_provider, marital_status, smoking_status, year,WEIGHT, STRATA, PSU)

nrow(analytic_type)
```


# diagnostic checks for cancer type availability


```{r}

# Summarize missingness across key analytic variables
# For cancer type, missingness is calculated only among respondents with a
# cancer history because cancer type is only defined in that group.
# For all other variables, missingness is calculated across the pooled dataset.


#(Supplementary table 3B)

tibble::tibble(
  Variable = c(
    "Cancer type (among cancer history = Yes only)",
    "Smoking status",
    "Insurance",
    "Marital status",
    "Personal health care provider",
    "Education",
    "Physical activity",
    "Race/Ethnicity",
    "Sex",
    "Age group",
    "Year",
    "WEIGHT",
    "STRATA",
    "PSU"
  ),
  Total_n = nrow(BRFSS_t),
  N_missing = c(
    sum(BRFSS_t$cancer_history == "Yes" & is.na(BRFSS_t$cancer_type), na.rm = TRUE),

    sum(is.na(BRFSS_t$smoking_status)),
    sum(is.na(BRFSS_t$insurance)),
    sum(is.na(BRFSS_t$marital_status)),
    sum(is.na(BRFSS_t$PHC_provider)),
    sum(is.na(BRFSS_t$education)),
    sum(is.na(BRFSS_t$physical_activity)),
    sum(is.na(BRFSS_t$race_ethnicity)),
    sum(is.na(BRFSS_t$sex)),
    sum(is.na(BRFSS_t$age_categorical)),
    sum(is.na(BRFSS_t$year)),
    sum(is.na(BRFSS_t$WEIGHT)),
    sum(is.na(BRFSS_t$STRATA)),
    sum(is.na(BRFSS_t$PSU))
  )
) %>%
  dplyr::mutate(
    pct_missing = round(100 * N_missing / Total_n, 2)
  ) %>%
  dplyr::arrange(dplyr::desc(pct_missing))
```


```{r}

# Missing cancer type by cancer history

BRFSS_t %>%
  dplyr::count(cancer_history, is.na(cancer_type), name = "n") %>%
  dplyr::arrange(cancer_history, desc(`is.na(cancer_type)`))
```


```{r}

# Missing cancer type by year/version/state

BRFSS_t %>%
  dplyr::filter(cancer_history == "Yes", is.na(cancer_type)) %>%
  dplyr::count(year, version, `_STATE`, sort = TRUE)
```


```{r}

# Distribution of cancer types among respondents with non-missing cancer type 
# (included in table one)

BRFSS_t %>%
  dplyr::filter(cancer_history == "Yes", !is.na(cancer_type)) %>%
  dplyr::count(cancer_type, name = "n") %>%
  dplyr::mutate(
    pct = round(100 * n / sum(n), 1),
    `Cancer Type` = as.character(cancer_type)
  ) %>%
  dplyr::select(`Cancer Type`, n, pct) %>%
  dplyr::arrange(match(`Cancer Type`, c("Breast cancer", "Prostate cancer", "Colon cancer", "Other cancer"))) %>%
  tibble::as_tibble()
```


```{r}

# missing cancer type by year/version/state

BRFSS_t %>%
  dplyr::filter(cancer_history == "Yes") %>%
  dplyr::count(cancer_type, name = "n") %>%
  dplyr::arrange(dplyr::desc(n))
```



```{r}

# Create survey design objects for cancer type-specific prevalence analyses
# Add factor-coded survey year, numeric year, and pooled year-specific PSU and
# strata identifiers for use in survey-weighted estimation across years.

analytic_type <- analytic_type %>%
  dplyr::mutate(
    year          = factor(year, levels = 2018:2024),
    year_num      = as.integer(as.character(year)),
    cancer_type   = factor(cancer_type,
                    levels = c("Breast cancer", "Colon cancer", "Prostate cancer", "Other cancer")),
    
    # Create unique strata IDs
    
    # BRFSS strata and PSU codes are assigned independently each year, so the same
    # numeric code can appear across years but refer to different design elements.
    # Interacting year with STRATA and PSU creates unique identifiers that prevent
    # the survey package from incorrectly treating codes from different years as the
    # same stratum or PSU in the pooled design.
    
    unique_strata = interaction(year, STRATA),
    unique_psu    = interaction(year, PSU)
  )

options(survey.lonely.psu = "adjust")

des_all <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = analytic_type,
  nest    = TRUE
)

# Cancer-type designs (with cancer history)
des_breast   <- subset(des_all, cancer_history == "Yes" & cancer_type == "Breast cancer"   & sex == "Female")
des_prostate <- subset(des_all, cancer_history == "Yes" & cancer_type == "Prostate cancer" & sex == "Male")
des_colon    <- subset(des_all, cancer_history == "Yes" & cancer_type == "Colon cancer")
des_other    <- subset(des_all, cancer_history == "Yes" & cancer_type == "Other cancer")

# Reference designs (no cancer history)
des_ref_female <- subset(des_all, cancer_history == "No" & sex == "Female")
des_ref_male   <- subset(des_all, cancer_history == "No" & sex == "Male")
des_ref_all    <- subset(des_all, cancer_history == "No")
```


```{r}

# Estimate annual weighted prevalence of physical activity by cancer type
# Calculate annual survey-weighted prevalence and 95% confidence intervals for
# each cancer type and for the corresponding no-cancer reference groups.


# Weighted prevalence by year for each cancer type
prev_breast <- survey::svyby(~physical_activity, ~year, des_breast,   
                              survey::svymean, vartype = "ci", na.rm = TRUE)
prev_prost  <- survey::svyby(~physical_activity, ~year, des_prostate, 
                              survey::svymean, vartype = "ci", na.rm = TRUE)
prev_colon  <- survey::svyby(~physical_activity, ~year, des_colon,    
                              survey::svymean, vartype = "ci", na.rm = TRUE)
prev_other  <- survey::svyby(~physical_activity, ~year, des_other,    
                              survey::svymean, vartype = "ci", na.rm = TRUE)

# Reference group prevalences
prev_ref_female <- survey::svyby(~physical_activity, ~year, des_ref_female, 
                                  survey::svymean, vartype = "ci", na.rm = TRUE)
prev_ref_male   <- survey::svyby(~physical_activity, ~year, des_ref_male,   
                                  survey::svymean, vartype = "ci", na.rm = TRUE)
prev_ref_all    <- survey::svyby(~physical_activity, ~year, des_ref_all,    
                                  survey::svymean, vartype = "ci", na.rm = TRUE)

```


```{r}

# This object stacks the yearly prevalence estimates for breast, prostate,
# colon, and other cancers for summary and reporting.


# Included inn supplementary table 4.

prev_all_types <- dplyr::bind_rows(
  tibble::as_tibble(prev_breast) %>% dplyr::mutate(cancer_type = "Breast cancer"),
  tibble::as_tibble(prev_prost)  %>% dplyr::mutate(cancer_type = "Prostate cancer"),
  tibble::as_tibble(prev_colon)  %>% dplyr::mutate(cancer_type = "Colon cancer"),
  tibble::as_tibble(prev_other)  %>% dplyr::mutate(cancer_type = "Other cancer")) %>%
  dplyr::select(cancer_type, year, physical_activity, ci_l, ci_u)

prev_all_types
```




```{r}

# Estimate annual percentage change (APC) by cancer type
# Add numeric survey year to each annual prevalence object, fit log-linear models
# to the annual weighted prevalence estimates, and convert the year coefficient
# to APC with 95% confidence intervals.

# Output included in table 2.

# Add year_num to each prevalence object
add_year_num <- function(df) {
  df %>% dplyr::mutate(year_num = as.integer(as.character(year)))
}

prev_breast <- add_year_num(prev_breast)
prev_prost  <- add_year_num(prev_prost)
prev_colon  <- add_year_num(prev_colon)
prev_other  <- add_year_num(prev_other)

# Log-linear APC models on annual weighted prevalences
lm_breast   <- lm(log(physical_activity) ~ year_num, data = prev_breast)
lm_prostate <- lm(log(physical_activity) ~ year_num, data = prev_prost)
lm_colon    <- lm(log(physical_activity) ~ year_num, data = prev_colon)
lm_other    <- lm(log(physical_activity) ~ year_num, data = prev_other)

# Extract APC with 95% CIs
get_apc <- function(fit, label) {
  broom::tidy(fit, conf.int = TRUE) %>%
    dplyr::filter(term == "year_num") %>%
    dplyr::transmute(
      group       = label,
      APC_pct     = round((exp(estimate) - 1) * 100, 2),
      ci_low_pct  = round((exp(conf.low) - 1) * 100, 2),
      ci_high_pct = round((exp(conf.high) - 1) * 100, 2),
      p_value     = p.value
    )
}

apc_type_results <- dplyr::bind_rows(
  get_apc(lm_breast,   "Breast cancer"),
  get_apc(lm_prostate, "Prostate cancer"),
  get_apc(lm_colon,    "Colon cancer"),
  get_apc(lm_other,    "Other cancer")
)

apc_type_results
```


Create analytic datasets for adjusted prevalence difference analyses

```{r}

# For each cancer type, create a comparison dataset that includes:
# - adults without a cancer history as the reference group
# - adults with the specified cancer type as the comparison group
# Breast cancer is restricted to females and prostate cancer is restricted to males.

dat_breast_pd <- analytic_type %>%
  dplyr::filter(
    sex == "Female",
    cancer_history == "No" | (cancer_history == "Yes" & cancer_type == "Breast cancer")
  ) %>%
  dplyr::mutate(
    group = factor(
      ifelse(cancer_history == "No", "No cancer history", "Breast cancer"),
      levels = c("No cancer history", "Breast cancer")
    )
  )

dat_prostate_pd <- analytic_type %>%
  dplyr::filter(
    sex == "Male",
    cancer_history == "No" | (cancer_history == "Yes" & cancer_type == "Prostate cancer")
  ) %>%
  dplyr::mutate(
    group = factor(
      ifelse(cancer_history == "No", "No cancer history", "Prostate cancer"),
      levels = c("No cancer history", "Prostate cancer")
    )
  )

dat_colon_pd <- analytic_type %>%
  dplyr::filter(
    cancer_history == "No" | (cancer_history == "Yes" & cancer_type == "Colon cancer")
  ) %>%
  dplyr::mutate(
    group = factor(
      ifelse(cancer_history == "No", "No cancer history", "Colon cancer"),
      levels = c("No cancer history", "Colon cancer")
    )
  )

dat_other_pd <- analytic_type %>%
  dplyr::filter(
    cancer_history == "No" | (cancer_history == "Yes" & cancer_type == "Other cancer")
  ) %>%
  dplyr::mutate(
    group = factor(
      ifelse(cancer_history == "No", "No cancer history", "Other cancer"),
      levels = c("No cancer history", "Other cancer")
    )
  )
```


```{r}

# Create survey design objects for adjusted prevalence difference analyses
# Construct survey design objects for each cancer type-specific comparison
# using the year-specific PSU and strata identifiers.

options(survey.lonely.psu = "adjust")

des_breast_pd <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = dat_breast_pd,
  nest    = TRUE
)

des_prostate_pd <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = dat_prostate_pd,
  nest    = TRUE
)

des_colon_pd <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = dat_colon_pd,
  nest    = TRUE
)

des_other_pd <- survey::svydesign(
  ids     = ~unique_psu,
  strata  = ~unique_strata,
  weights = ~WEIGHT,
  data    = dat_other_pd,
  nest    = TRUE
)
```


```{r}

# Fit adjusted linear probability models for each cancer type comparison
# These models estimate adjusted prevalence differences in physical activity
# between each cancer type group and adults without a cancer history.
# Sex is included only in the colon and other-cancer models because those
# comparisons include both males and females

m_breast <- survey::svyglm(
  physical_activity ~ group + year + age_categorical + race_ethnicity +
    education + insurance + PHC_provider + marital_status + smoking_status,
  design = des_breast_pd,
  family = gaussian(link = "identity")
)

m_prostate <- survey::svyglm(
  physical_activity ~ group + year + age_categorical + race_ethnicity +
    education + insurance + PHC_provider + marital_status + smoking_status,
  design = des_prostate_pd,
  family = gaussian(link = "identity")
)

m_colon <- survey::svyglm(
  physical_activity ~ group + year + sex + age_categorical + race_ethnicity +
    education + insurance + PHC_provider + marital_status + smoking_status,
  design = des_colon_pd,
  family = gaussian(link = "identity")
)

m_other <- survey::svyglm(
  physical_activity ~ group + year + sex + age_categorical + race_ethnicity +
    education + insurance + PHC_provider + marital_status + smoking_status,
  design = des_other_pd,
  family = gaussian(link = "identity")
)
```


```{r}

# Extract adjusted prevalence difference estimates
# Extract the adjusted prevalence difference for each cancer type comparison
# and report results in percentage points with 95% confidence intervals


# Output included in supplementary table 5.

get_apd <- function(fit, label) {
  broom::tidy(fit, conf.int = TRUE) %>%
    dplyr::filter(grepl("group", term)) %>%
    dplyr::transmute(
      comparison = label,
      aPD_pp     = round(estimate * 100, 1),
      ci_low_pp  = round(conf.low * 100, 1),
      ci_high_pp = round(conf.high * 100, 1),
      p_value    = p.value
    )
}

apd_type_results <- dplyr::bind_rows(
  get_apd(m_breast,   "Breast cancer"),
  get_apd(m_prostate, "Prostate cancer"),
  get_apd(m_colon,    "Colon cancer"),
  get_apd(m_other,    "Other cancer")
)

apd_type_results
```


```{r}

# Prepare weighted prevalence estimates for plotting
# Convert each prevalence object into a consistent data frame with year,
# group label, prevalence estimate, and confidence interval limits.

# Helper function to prepare prevalence data for plotting
prep_prev <- function(prev_obj, label) {
  tibble::as_tibble(prev_obj) %>%
    dplyr::transmute(
      year    = as.integer(as.character(year)),
      group   = label,
      prev    = physical_activity,
      ci_low  = ci_l,
      ci_high = ci_u
    )
}

# Combine cancer type and reference group for each plot

# Combine each cancer type with its corresponding reference group
# Breast cancer is compared with females without a cancer history.
# Prostate cancer is compared with males without a cancer history.
# Colon cancer and other cancers are compared with all adults without a cancer history.

plot_breast_df   <- dplyr::bind_rows(prep_prev(prev_breast, "Breast cancer"),   prep_prev(prev_ref_female, "No cancer history"))
plot_prostate_df <- dplyr::bind_rows(prep_prev(prev_prost,  "Prostate cancer"),  prep_prev(prev_ref_male,   "No cancer history"))
plot_colon_df    <- dplyr::bind_rows(prep_prev(prev_colon,  "Colon cancer"),     prep_prev(prev_ref_all,    "No cancer history"))
plot_other_df    <- dplyr::bind_rows(prep_prev(prev_other,  "Other cancers"),     prep_prev(prev_ref_all,    "No cancer history"))


# Define plotting function
# This function plots annual weighted prevalence of physical activity with
# 95% confidence intervals for the cancer type group and the reference group.

plot_prev <- function(df) {
  ggplot(df, aes(x = year, y = prev * 100, group = group)) +
    geom_errorbar(aes(ymin = ci_low * 100, ymax = ci_high * 100),
                  width = 0.15, linewidth = 0.45) +
    geom_line(aes(linetype = group), linewidth = 0.9) +
    geom_point(aes(shape = group), size = 2.5, fill = "white", stroke = 0.8) +
    scale_x_continuous(breaks = 2018:2024) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                       expand = expansion(mult = c(0, 0.02))) +
    labs(x = "Survey year",
         y = "Weighted prevalence of PA (%)",
         linetype = NULL, shape = NULL) +
    theme_classic(base_size = 12) +
    theme(
      axis.text       = element_text(color = "black"),
      legend.text      = element_text(size = 14),
      legend.key.width = unit(2, "lines"),
      legend.position = "top"
    )
}

# Generate plots
p_breast   <- plot_prev(plot_breast_df)
p_prostate <- plot_prev(plot_prostate_df)
p_colon    <- plot_prev(plot_colon_df)
p_other    <- plot_prev(plot_other_df)

# Save
ggsave("Fig_Breast_vs_NoCancer.png",   p_breast,   width = 6.5, height = 4.5, units = "in", dpi = 1200)
ggsave("Fig_Prostate_vs_NoCancer.png", p_prostate, width = 6.5, height = 4.5, units = "in", dpi = 1200)
ggsave("Fig_Colon_vs_NoCancer.png",    p_colon,    width = 6.5, height = 4.5, units = "in", dpi = 1200)
ggsave("Fig_Other_vs_NoCancer.png",    p_other,    width = 6.5, height = 4.5, units = "in", dpi = 1200)
```


```{r}

# These figures are included in the manuscript as Supplemental Figure 2.

p_breast  
p_prostate 
p_colon
p_other
```


# Supplemental data checks and flow-diagram counts

```{r}

# Sample derivation for the cancer type analytic dataset

# Summarize sequential exclusions used to create the complete-case analytic dataset
# for cancer type comparisons.

tibble::tibble(
  step = c(
    "Adults in pooled BRFSS dataset from states administering the optional cancer type module, 2018–2024",
    "Excluded: respondents with a cancer history but missing cancer type",
    "Excluded: missing physical activity, covariates, or survey design variables",
    "Final analytic dataset"
  ),
  n = c(
    nrow(BRFSS_t),
    nrow(BRFSS_t) -
      nrow(BRFSS_t %>% dplyr::filter(!(cancer_history == "Yes" & is.na(cancer_type)))),
    nrow(BRFSS_t %>% dplyr::filter(!(cancer_history == "Yes" & is.na(cancer_type)))) -
      nrow(
        BRFSS_t %>%
          dplyr::filter(!(cancer_history == "Yes" & is.na(cancer_type))) %>%
          tidyr::drop_na(
            physical_activity, age_categorical, sex, race_ethnicity, education,
            insurance, PHC_provider, marital_status, smoking_status,
            year, WEIGHT, STRATA, PSU
          )
      ),
    nrow(analytic_type)
  )
) %>%
  knitr::kable(caption = "Sample derivation for the cancer type analytic dataset")
```




